Bootstrap learning via modular concept discovery

### 2013-02-01 drafts/ijcai2013/ijcai13.tex
Changes

x\documentclass{article}

@@

\section{Introduction}
Many of the problems we want to solve in AI are best cast as search
problems. There are many search algorithms, but they typically exploit
local or topological structure in the search space. For example,
convex optimization and gradient based search methods~\cite{} take
advantage of local smoothness to move in good directions. Local search
algorithms move from one candidate solution to a neighboring
one\cite{}. Heuristic search techniques take advantage of topological
structure to prune off large parts of the solution space~\cite{}.

However, many of the problems we want solve -- discovering scientific
theories, developing new technologies, figuring out how to bake -- do
not come with such a structured search space. This is particularly
true of problems, as these are, whose solutions can be naturally
framed as programs. Here we ask how to extract structure from the
space of programs by attempting to solve multiple related problems in
a given domain at once.

We take as inspiration the programmer who -- having reused several
motifs, programming patterns, or subroutines to solve different but
similar simple tasks -- creates a function library to help manage the
complexity of larger projects. Our algorithm builds a library
of reusable program components and a distribution over these,
effectively learning from simple tasks a search space structure which
enables it to solve more complex tasks.

@@

Concretely, our algorithm works as follows. It begins with a small
library of program primitives and a distribution over programs
synthesized from this library. In each iteration, it generates
programs in order of highest probability according to the library
learned in the previous iteration. From the discovered solutions, it
finds new reusable functions that compress these solutions, and it
re-estimates the distribution over programs by weighting these new
functions according to their frequencies. In using compression as a
guide to the choice representation, we instantiate the idea that good
representations are those which minimize the description length of
typical elements in the domain.

@@

\section{Related Work}
The idea of discovering and using reusable subcomponents is an old one
in AI. Rendell's 1985 paper~\cite{DBLP:conf/ijcai/Rendell85},
``Substantial Constructive Induction Using Layered Information
Compression: Tractable Feature Formation in Search,'' contained many
of these ideas. It did not, however, use programs as its solution
space representation. Within the Inductive Logic Programming
literature, researchers have explored ``Constructive Induction,''
generating new and reusable logic programming
terms~\cite{DBLP:conf/ijcai/Muggleton87}. We believe that applying
constructive induction to a compositional functional programming
paradigm will provide even greater benefits.

Dicovering useful subroutines has been explored in Genetic
Programming as
well~\cite{DBLP:conf/foga/Koza92}~\cite{DBLP:conf/ices/KozaBAK96}.

The importance of a curriculum has also been explored previously,
though generally the curriculum has been explicitly given rather than
discovered. Quinlan and Cameron-Jones~\cite{Quinlan93foil:a} showed
that a logic program learning algorithm could learn sequentially more
difficult list programs if the algorithm was provided with its own
solutions to previously solved simple problems. More recently, several
researchers have observed that presenting problems to machine learning
algorithms in order of difficulty results in better
performance~\cite{Krueger2009380}~\cite{DBLP:conf/icml/BengioLCW09}.

Our use of a stochastic grammar over combinatory logic is based on
Liang et al.~\cite{liang10programs}. In that work, a non-parametric
distribution was placed over combinatory logic expressions, and an
MCMC algorithm was developed to perform joint inference over these
expressions and the underlying stochastic grammar. Our approach is
inspired by this generative model, but avoids the difficulties
associated with a stochastic local search over programs (the same
difficulties faced by genetic programming), especially when the
likelihood function (or the fitness function) are not of much help.

@@

The remainder of this paper will include two major sections.  We will
first describe the representation language we chose and the details of
the exploration and compression steps of the algorithm. We will then
present results from two experiments involving the domains of
symbolic regression and boolean circuit learning.

@@

\section{Extracting modular concepts from functional programs}

Our goal is to solve the following minimal feedback multi-task program
induction problem: suppose we are given set of \textbf{tasks}
$T=\{t_k\}_{k=1}^K$ where each task is a function $t_i : \mathcal{L}
\rightarrow \{0, 1\}$ and where $\mathcal{L}$ is the set of programs
in our language.  We say that $ p \in \mathcal{L}$ \textbf{solves}
$t_i$ if $ t_i(p) = 1$. Our goal is to solve as many of the tasks in
$T$ as we can.

We propose the \textbf{E-C algorithm}, an iterative procedure to solve this
problem. It will be useful in presenting the algorithm to define the
following terminology: a \textbf{frontier} is the finite set of
programs that the algorithm considers in any one iteration. The
\textbf{frontier size} $N$ is the prespecified number of elements in
the frontier. We will say that a task is \textbf{hit} by the frontier
if there is a program in the frontier that solves it.

@@

What remains is to specify a representation language $\mathcal{L}$, a
form for the distribution $\mathcal{D}$, a compression criterion for a
set of expressions in $\mathcal{L}$. We also need to describe
procedures for the enumeration of the frontier, compression of the
expressions, and re-estimations of $\mathcal{D}$.

@@

Following~\cite{liang10programs} and~\cite{Briggs:2008}, we use a
simply typed combinatory logic -- a variable-free subset of the
polymorphic simply-typed lambda calculus~\cite{Pierce_2002} -- as our
program representation language. In short, the combinatory logic
introduces a \emph{basis} of several primitive \emph{combinators} such
that any function in the lambda calculus can be alternatively written
as applications of the basis combinators. Some common basis
combinators are defined as follows:
\begin{align}
I\; x &\rightarrow  x \\
S\; f\; g\; x\; &\rightarrow (f\; x)\; (g\; x) \\ 
B\; f\; g\; x\; &\rightarrow (f\; x)\; g \\ 
C\; f\; g\; x\; &\rightarrow f\; (g\; x) \\ 
\end{align}

The basis combinators are theoretically enough to express any Turing
computable function, but we will assume that our language has a
variety of primitives and, together with the basis combinators, we
call these the primitive combinators. Note that by default we consider
all primitives to be in their ``curried'' form (e.g. a function like
``+'' adds two numbers $x$ and $y$ by being first applied to $x$,
returning the function $(+\;x)$ and then being applied to $y$,
returning the object $(+\;x)\;y$).

The basis combinators can themselves been expressed in the lambda
calculus (with definitions following directly from the equations
above). The lambda calculus has two basic operations -- application
and abstraction -- but in using the combinatory logic we sequester
uses of the abstraction operation inside the combinators. In doing so,
we have replaced the variable binding of the $\lambda$ operator with
the variable \emph{routing} of these basis combinators. Our
representation thus becomes variable-free. See Liang et
al.~\cite{liang10programs} for a more detailed discussion of this
routing interpretation.

Using combinatory logic is very convenient for program synthesis:
since every expression is the application of one expression to another
-- with this recursion bottoming out at the primitive combinators --
each program is a binary tree. Most importantly, any subtree is itself
a well-formed expression; this is not the case in the lambda calculus,
since abstraction introduces long range dependencies between the
$\lambda$ operator and the variables to which that operator refers. In
the lambda calculus, then, a subtree might have free variables, not
bound to any enclosing $\lambda$.

As a simple example of our representation, consider the squaring
function $\lambda x. * x\; x $. Using two of the basis combinators
above, we can write the squaring function in combinatory logic as $S *
I$ (taking $*$ as a primitive), where application associates to the
left, so we drop the parentheses. When we apply this combinator to a
value $x$, the action of the combinator is defined as $S \;*\; I\;
x~\rightarrow~(*\; x)\; (I\; x)~\rightarrow(*\; x)\; x$.

We can extend this representation with a simple polymorphic type
system~\cite{Pierce_2002}. In this representation, a type is either a
type primitive (e.g. reals, integers, booleans, etc.), a type variable
(e.g. $\sigma$), or a function type $ t_1 \rightarrow t_2$ of
functions from source type $t_1$ to target type $t_2$. Any combinator
can be represented as a binary tree~(Figure~\ref{fig:clbintree}) whose
leaves are typed primitive combinators and whose interior nodes
represent typed applications of combinators (we annotate interior
nodes with their types).

\begin{figure}
%%  \centering
  \begin{subfigure}[]{0.45\linewidth}
    \begin{tikzpicture}[scale=.8,
        transform shape, level/.style={sibling distance=5mm/#1, font=\scriptsize}
        ]
      \node [font=\scriptsize] (top) {$\mathbb{R}$}
      child {node [above=.3cm, left=.1cm, text width=1.8cm] (a) {$(\mathbb{R} \rightarrow \mathbb{R}) \rightarrow \mathbb{R} 
          \rightarrow \mathbb{R}$}
        child {node [left=.5cm, below=1cm, text width=3.0cm] 
          {$ \mathbf{S} : (t_2 \rightarrow t_1 \rightarrow t_0)  \rightarrow (t_2 \rightarrow t_1) 
            \rightarrow \\ t_2 \rightarrow t_0$}}
        child {node [right=0cm, text width=2cm] {$ \mathbf{*}: \mathbb{R} \rightarrow \mathbb{R} \rightarrow \mathbb{R}$}}
      }
      child {node [below right=-.5cm and .2cm] (b) {$\mathbb{R} \rightarrow \mathbb{R}$}
        child {node [above=.5cm] {$ \mathbf{I}: t_0 \rightarrow t_0 \rightarrow t_0$}}
      }
    \end{tikzpicture}
    \caption{}
    \label{fig:clbintree}
  \end{subfigure}
  \begin{subfigure}[R]{0.5\linewidth}
    \begin{tikzpicture}[scale=.7, transform shape, 
        level/.style={sibling distance=10mm/#1},
        every edge/.style={draw, thin}
      ]
      \node [circle,draw] (a) {}
      child {node [left=0cm] (b) {$f_1$}}
      child {node [left=.2cm] (c) {\dots} edge from parent[draw=none] }
      child {node [left=.4cm] (e) {$f_n$}}
      child {node [circle,draw, left=1cm] (d) {} edge from parent[very thick]
        child {node [circle, draw, left=.25cm] (l) {} edge from parent[draw=none]
          child {node [left=.5cm] (b) {$f_1$}} 
          child {node [left=.3cm] (c) {\dots} edge from parent[draw=none] }
          child {node [left=.3cm] (e) {$f_n$} edge from parent[very thick]}
          child {node [circle, draw, left=.25cm] (and2) {}
            child {node [circle, draw, left=.25cm](l2) {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
            child {node [circle, draw] (r2)  {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
          }
        }
        child {node [circle, draw, right=.25cm] (r) {} edge from parent[draw=none]
          child {node [circle, draw, right=.25cm] (and3) {}
            child {node [circle, draw ](l3) {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
            child {node [circle, draw, right=.2cm ] (r3)  {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
          }
          child {node [right=.3cm] (b) {$f_1$} edge from parent[very thick]} 
          child {node [right=.4cm] (c) {\dots} edge from parent[draw=none] }
          child {node [right=.6cm] (e) {$f_n$}}
        }
      };

@@

      \path (and3)  edge node[inner sep=0mm,pos=.4] (t1) {} node {} (l3);
      \path (and3)  edge node[inner sep=0mm,pos=.4] (k1) {} node {} (r3);
      \path (t1) edge[bend right] (k1) ;
    \end{tikzpicture}
    \caption{AND/OR expression search space. }
    \label{fig:andor}
    \end{subfigure}
  \caption{(a) The typed combinator $S * I$ ($f(x)=x^2$) represented as
    binary tree. (b) The space of all typed combinators represented as an
    infinitely deep AND/OR tree. Any specific combinator is a
    sub-binary tree of this AND/OR tree whose leaves consist of only
    the $f_i$ nodes (i.e., the primitive combinators of the grammar).}

@@

We specify a distribution $\mathcal{D}$ over expressions in
$\mathcal{L}$ as a stochastic
grammar~\cite{feldman1969grammatical}. Many stochastic grammars have
the desirable property that the probability of an expression is the
product of the probabilities of its subcomponents. The distribution we
specify will be a particularly simple verson of this, in which the
probability of an expression is the product of the probability of its
leaves.

$\mathcal{D}$ will associate with each primitive combinator $c_i$ a
prior probabilities $p_i$, $0~\leq~p_i~\leq~1, \sum_i~p_i~=~1 $. The
probability of a leaf withr primitive combinator $c_i$ will depend on
which other primitive combinators could have been chosen in its
place. This set will depend on the type that was requested by the
parent of this leaf when the leaf's primitive was chosen; we call this
type the \emph{requesting} type of the leaf. For example, suppose that
we have the primitive $(+1)$ of type $Int \rightarrow Int$ and the
primitive $NOT$ of type $Bool \rightarrow Bool$. If the requesting
type is $\sigma \rightarrow Int$, then we can only use $(+1)$, but if
the requesting type is $\sigma \rightarrow \sigma$ then we could use
either $(+1)$ or $NOT$. Let $C = c_1, \dots, c_n$ be the set of
primitive combinators in our grammar; we define $C_t$ for any type
$\tau$ to be the set of combinators in $C$ that are consistent with
$\tau$, that is, whose types \emph{unify} with $\tau$ (we use
unification according to the standard type theory
presentation~\ref{Pierce_2002}).

Given a distribution $\mathcal{D} = (p_1, \dots, p_n)$, we define the
probability of an expression $e \in \mathcal{L}$ to be
$p(e)~=~\prod_{n=1}^N p(c_n | \tau(c_n))$ where $p(c_n | \tau(c_n))$ is the
probability of using primitive $c_n$ when the requesting type is
$\tau(c_n)$. In turn, we define the conditional probabilities to be
\[
p(c_i | \tau(c_n)) \propto \frac{p_i}{\sum_{C_{\tau(c_n)}} p_j}.
\]
That is, $p(c_i | \tau(c_n)$ is proportional to the probability of
sampling $c_i$ from the multinomial distribution defined by the
$p_i$'s, conditioned on the requesting type, with the constant of
proportionality chosen to ensure that the sum over the probabilities
of all expressions converges to a finite value (so that the
distribution over expressions is proper). For binary trees, this is
true whenever the constant of proportionality is less than
$\frac{1}{4}$\footnote{The probability mass $M_d$ of all expressions
  of depth less than or equal to $d$ can be written with the
  recurrence relation $M_d \leq M_{d-1}^2 + M_1$. This has an
  asymptotic fixed point, as $d \rightarrow \infty$ if $x = x^2 + M_1$
  has a solution, which is true if $M_1 \leq 1/4$.}.

Now, suppose that we have a set of expressions $\mathcal{E}$, and we
wish to estimate the maximum likelihood values of the $p_i$'s. If the
choice of primitive at each leaf were not conditioned on the
requesting type, we could just count all occurences of each primitive
in our expression set, and this would be proportional to the maximum
likelihood estimate. This is ML estimator for a multinomial
distribution. However, the fact that we condition on the requesting
type makes this a considerably more difficult optimization task. One
of the simplest approximations we can make is to calculate the
frequency with which each primitive combinator appears in
$\mathcal{E}$ when the requesting type is such that it could have
been chosen. This is a particularly straightforward calculuation, and
we use it for its convenience, finding that our empirical results
justify its use. Future inquiry will be needed to determine to what
extent a more accurate estimator can improve these results.

\subsection{Best-first enumeration of programs}
In order to explore the frontier of most promising programs, we need a
procedure that enumerates the $N$ most probable programs. There has
been recent interest in this problem, most of it focused on
enumerating the shortest program satisfying some
criterion~\cite{DBLP:conf/sfp/Katayama05}
\cite{DBLP:conf/aaip/YakushevJ09}. Our approach is to formulate the
problem as a best-first exploration of an AND/OR tree
(Figure~\ref{fig:andor})~\cite{nilsson1982principles}~\cite{DBLP:journals/cacm/Hall73},
which we denote $\mathcal{G}$. In words, the enumeration procedure is
best described by the following recursion: every program is either a
primitive combinator or it is a left child program applied to a right
child program.

This can framed as the following AND/OR tree. Suppose we want a
program of type $t$. The root of $\mathcal{G}$ is an OR node of type
$t$. Its children are elements in $C_t$ (defined in the previous
section) and one AND node of type $t$ with two children. Each child of
an AND node has the same structure as the root node, with modified
types: type of its left child is $\tau \rightarrow t$ (where $\tau$ is
a fresh type variable). Since the type of the right child is
constrained by the subtree rooted at the left child, we always expand
the left child first. Once we have a complete left child program, we
can use its target type as the type of the right child.

Recall that a valid partial solution $\mathcal{H}$ is a subtree of
$\mathcal{G}$ satisfying the following properties: the root of
$\mathcal{H}$ is the root $\mathcal{G}$ and any OR node in
$\mathcal{H}$ has at most one child. $\mathcal{H}$ is a complete
solution (and thus a complete program) if it is a partial solution
whose leaves are leaves of $\mathcal{G}$, if each OR node in
$\mathcal{H}$ has exactly one child, and if each AND node in
$\mathcal{H}$ is connected to all of that node's children in
$\mathcal{G}$

@@

\subsection{Re-estimating the stochastic grammar over programs.}

@@

In the formulation we consider, each task $t$ corresponds to a
polynomial $f$, and an expression $e$ solves $t$ if ${ (e\;
  i)=f(i) }$ for all ${ i=0,\dots,9 }$.

Each task is to predict the value of an unknown polynomial on the
inputs $0, \dots, 9$. The set of tasks corresponds to the set of
polynomials $\{ax^2 + bx + c | a, b, c \in \{0, \dots, 9 \}\}$. In the
initial library, we include the basic combinators I, S, B, C and four
arithmetic primitives, $1$, $0$, $*$, and $+$. The choice of initial
weights on these primitives can make a difference, as it determines
the relative ordering of combinators in the enumeration step. To get a
general sense of the algorithm's performance, we set the initial
weights to be slightly different on each run, perturbing them around a
uniform weighting.

\begin{figure}[h]
\centering
\begin{subfigure}[Before]{\linewidth}
\includegraphics[width=.9\linewidth]{figures/symreg3frontiers.pdf}
\label{fig:symregLearning}
\caption{Learning curves as a function of frontier size.  As frontier
  size is increased, it plateaus closer to $100\%$ performance. A
  control brute-force search over $150000$ only hits $3\%$ of the
  tasks.}
\end{subfigure}

>>>>

\documentclass{article}

@@

\section{Introduction}
Many of the problems we want to solve in AI are best cast as search
problems. There are many search algorithms, but they typically exploit
local or topological structure in the search space. For example,
convex optimization and gradient based search methods take
advantage of local smoothness to move in good directions. Local search
algorithms move from one candidate solution to a neighboring
one. Heuristic search techniques take advantage of topological
structure to prune off large parts of the solution space.

However, many of the problems we want solve -- discovering scientific
theories, developing new technologies, figuring out how to bake -- do
not come with such a structured search space. This is particularly
true of problems, such as those mentioned, whose solutions can be
naturally framed as programs. Here we ask how to extract structure
from the space of programs by attempting to solve multiple related
problems in a given domain at once.

We take as inspiration the programmer who -- having reused several
subroutines to solve different but similar simple tasks -- creates a
function library to help manage the complexity of larger projects. The
EC algorithm, which we introduce here, builds a library of reusable
program components and places a distribution over these, effectively
learning from simple tasks a search space structure that enables it
to solve more complex tasks.

@@

Concretely, the EC algorithm works as follows. It begins with a small
library of program primitives and a distribution over programs
synthesized from this library. In each iteration, it generates
programs in order of highest probability according to the library
learned in the previous iteration. From the discovered solutions, it
finds new reusable functions that compress these solutions, and it
re-estimates the distribution over programs by weighting these new
functions according to their frequencies. In using compression as a
guide to the choice of representation, we instantiate the idea that
good representations are those which minimize the description length
of typical elements in the domain.

@@

\section{Related Work}
The idea of discovering and using reusable subcomponents is an old one
in AI. Rendell's 1985 paper~\cite{DBLP:conf/ijcai/Rendell85},
``Substantial Constructive Induction Using Layered Information
Compression: Tractable Feature Formation in Search,'' presented within
a classical search paradigm the idea that compression provides a
heuristic for constructing good representations. Within the Inductive
Logic Programming (ILP) literature, researchers have explored ``Constructive
Induction,'' generating new and reusable logic programming
terms~\cite{DBLP:conf/ijcai/Muggleton87}. We apply these ideas to
learning functional programs, and believe that the modularity and
compositionality afforded by this representation is necessary for
these ideas to be successful.

Genetic programming
(GP)~\cite{DBLP:conf/foga/Koza92}~\cite{DBLP:conf/ices/KozaBAK96} has
tackled the problem of program learning and has explored the notion
that reusable functions are helpful. This work relies, however, on
stochastic local search over the space of programs, and automatically
discovering useful subroutines is seen as a helpful heuristic. More
recently, Liang et al.~\cite{liang10programs} used a stochastic
grammar over combinatory logic to induce shared structure in
multi-task program induction problems. We find this latter
representation compelling and use it here, but we see both this and
genetic programming as tackling the question of how to benefit from
shared structure in situations where local search might be successful
on its own. Such problems rely on domain knowledge -- whether in the
form of a fitness function or a likelihood function -- to provide
local structure to the search space.

But thes problems we want AI systems to solve often lack this kind of
built in structure. These are the kinds of problems more often
associated with ILP and classical AI than with GP and machine
learning. 
\\

@@

The remainder of this paper will include two major sections. In the
first, we describe in detail the EC algorithm and design choices we
make in implementing it. We will then present results from two
experiments in the domains of symbolic regression and Boolean
functioin learning.

@@

\section{Extracting concepts from programs}

Our goal is to solve the following minimal feedback multi-task program
induction problem: suppose we are given a set of \textbf{tasks}
$T=\{t_k\}_{k=1}^K$ where each task is a function $t_k : \mathcal{L}
\rightarrow \{0, 1\}$ and where $\mathcal{L}$ is the set of
expressions in our language.  We say that expression $ e \in
\mathcal{L}$ \textbf{solves} $t_k$ if $ t_k(p) = 1$. Our goal is to
solve as many of the tasks in $T$ as we can.

We propose the \textbf{E-C algorithm}, an iterative procedure to solve this
problem. It will be useful in presenting the algorithm to define the
following terminology: a \textbf{frontier} is the finite set of
expressions that the algorithm considers in any one iteration. The
\textbf{frontier size} $N$ is the prespecified number of elements in
the frontier. We will say that a task is \textbf{hit} by the frontier
if there is a expression in the frontier that solves it.

@@

Next, we specify a representation language $\mathcal{L}$, a
form for the distribution $\mathcal{D}$, and a compression criterion for a
set of expressions in $\mathcal{L}$. We also describe
procedures for the enumeration of the frontier, compression of the
expressions, and re-estimation of $\mathcal{D}$.

@@

Following~\cite{liang10programs} and~\cite{Briggs:2008}, we use a
simply typed combinatory logic -- a variable-free subset of the
polymorphic simply-typed lambda calculus~\cite{Pierce_2002} -- as our
program representation language. In short, the combinatory logic
introduces a \emph{basis} of several primitive \emph{combinators} such
that any function can be written as applications of the basis
combinators. Some common basis combinators are defined as follows:
\begin{align}
\mathbf{I}\; x &\rightarrow  x & \text{ (identity) }\\
\mathbf{S}\; f\; g\; x\; &\rightarrow (f\; x)\; (g\; x)\\ 
\mathbf{C}\; f\; g\; x\; &\rightarrow (f\; x)\; g  \\ 
\mathbf{B}\; f\; g\; x\; &\rightarrow f\; (g\; x) & \text{ (composition) }
\end{align}

The basis combinators are theoretically enough to express any Turing
computable function, but we will assume that our language has a
variety of primitives and, together with the basis combinators, we
call these the \textbf{primitive combinators}. Note that by default we consider
all primitives to be in their ``curried'' form (e.g. a function like
``+'' adds two numbers $x$ and $y$ by being first applied to $x$,
returning the function $(+\;x)$ and then being applied to $y$,
returning $(+\,x\,y)$\;).

The basis combinators can themselves be expressed in the lambda
calculus (with definitions following directly from the equations
above). The lambda calculus has two basic operations -- application
and abstraction (using the $\lambda$ operator -- but in using the
combinatory logic we sequester uses of the abstraction operation
inside the combinators. In doing so, we have replaced the variable
binding of the $\lambda$ operator with the variable \emph{routing} of
these basis combinators. Our representation thus becomes
variable-free. See Liang et al.~\cite{liang10programs} for a more
detailed discussion of this routing interpretation.

Using combinatory logic is very convenient for program
synthesis~\cite{briggs2006functional}: since every expression is the
application of one expression to another -- with this recursion
bottoming out at the primitive combinators -- each program is a binary
tree. Most importantly, any subtree is itself a well-formed
expression; this is not the case in the lambda calculus, since
abstraction introduces long range dependencies between the $\lambda$
operator and the variables to which that operator refers. In the
lambda calculus, then, a subtree might have free variables, not bound
to any enclosing $\lambda$.

As a simple example of our representation, consider the squaring
function $\lambda x. * x\; x $. Using two of the basis combinators
above, we can write the squaring function in combinatory logic as $S\; *\;
I$ (taking $*$ as a primitive), where application associates to the
left, so we drop the parentheses. When we apply this combinator to a
value $x$, the action of the combinator is defined as $S \;*\; I\;
x~\rightarrow~(*\; x)\; (I\; x)~\rightarrow *\; x \; x$.

We can extend this representation with a simple polymorphic type
system~\cite{Pierce_2002}. In this representation, a type is either a
type primitive (e.g. reals, integers, booleans, etc.), a type variable
(e.g. $\sigma$), or a function type $ \tau_1 \rightarrow \tau_2$ of
functions from source type $\tau_1$ to target type $\tau_2$. Any combinator
can be represented as a binary tree~(Figure~\ref{fig:clbintree}) whose
leaves are typed primitive combinators and whose interior nodes
represent typed applications of combinators (we annotate interior
nodes with their types).

\begin{figure}
%%  \centering
  \begin{subfigure}[Before]{0.45\linewidth}
    \begin{tikzpicture}[scale=.8,
        transform shape, level/.style={sibling distance=5mm/#1, font=\scriptsize}
        ]
      \node [font=\scriptsize] (top) {$\mathbb{R}$}
      child {node [above=.3cm, left=.1cm, text width=1.8cm] (a) {$(\mathbb{R} \rightarrow \mathbb{R}) \rightarrow \mathbb{R} 
          \rightarrow \mathbb{R}$}
        child {node [left=.5cm, below=1cm, text width=3.0cm] 
          {$ \mathbf{S} : (t_2 \rightarrow t_1 \rightarrow t_0)  \rightarrow (t_2 \rightarrow t_1) 
            \rightarrow \\ t_2 \rightarrow t_0$}}
        child {node [right=0cm, text width=2cm] {$ \mathbf{*}: \mathbb{R} \rightarrow \mathbb{R} \rightarrow \mathbb{R}$}}
      }
      child {node [below right=-.5cm and .2cm] (b) {$\mathbb{R} \rightarrow \mathbb{R}$}
        child {node [above=.5cm] {$ \mathbf{I}: t_0 \rightarrow t_0 \rightarrow t_0$}}
      }
    \end{tikzpicture}
    \caption{    \label{fig:clbintree}}
  \end{subfigure}
  \begin{subfigure}[After]{0.5\linewidth}
    \begin{tikzpicture}[scale=.7, transform shape, 
        level/.style={sibling distance=10mm/#1},
        every edge/.style={draw, thin}
      ]
      \node [circle,draw] (a) {}
      child {node [left=0cm] (b) {$f_1$}}
      child {node [left=.2cm] (c) {\dots} edge from parent[draw=none] }
      child {node [left=.4cm] (e) {$f_n$}}
      child {node [circle,draw, left=1cm] (d) {} edge from parent[very thick]
        child {node [circle, draw, left=.25cm] (l) {} edge from parent[draw=none]
          child {node [left=.5cm] (b) {$f_1$}} 
          child {node [left=.3cm] (c) {\dots} edge from parent[draw=none] }
          child {node [left=.3cm] (e) {$f_n$} edge from parent[very thick]}
          child {node [circle, draw, left=.25cm] (and2) {}
            child {node [circle, draw, left=.25cm](l2) {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
            child {node [circle, draw] (r2)  {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
          }
        }
        child {node [circle, draw, right=.25cm] (r) {} edge from parent[draw=none]
          child {node [circle, draw, right=.25cm] (and3) {}
            child {node [circle, draw ](l3) {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
            child {node [circle, draw, right=.2cm ] (r3)  {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
          }
          child {node [right=.3cm] (b) {$f_1$} edge from parent[very thick]} 
          child {node [right=.4cm] (c) {\dots} edge from parent[draw=none] }
          child {node [right=.6cm] (e) {$f_n$}}
        }
      };

@@

      \path (and3)  edge node[inner sep=0mm,pos=.4] (t1) {} node {} (l3);
      \path (and3)  edge node[inner sep=0mm,pos=.4] (k1) {} node {} (r3);
      \path (t1) edge[bend right] (k1) ;
    \end{tikzpicture}
    \caption{  \label{fig:andor}}
    \end{subfigure}
  \caption{(a) The typed combinator $S * I$ ($f(x)=x^2$ over real values ($\mathbb{R}$)) represented as a
    binary tree. (b) The space of all typed combinators represented as an
    infinitely deep AND/OR tree.}

@@

We specify a distribution $\mathcal{D}$ over expressions in
$\mathcal{L}$ as a stochastic
grammar~\cite{feldman1969grammatical}. Many stochastic grammars have
the desirable property that the probability of an expression is the
product of the probabilities of its subcomponents. The distribution we
specify will be a simple verson of this, in which the
probability of an expression is the product of the probability of its
leaves.

Let ${C = c_1, \dots, c_N }$ be the primitive combinators in
$\mathcal{L}$. ${ \mathcal{D}=p_1, \dots, p_N }$ will associate with
each primitive combinator $c_i$ a prior probability $p_i$,
$0~\leq~p_i~\leq~1, \sum_i~p_i~=~1 $. The probability of a leaf with
primitive combinator $c_i$ will depend on which other primitive
combinators could have been chosen in its place. This set will depend
on the type that was requested by the parent of this leaf when the
leaf's primitive was chosen; we call this type the \emph{requesting}
type of the leaf. For example, suppose that we have the primitive
$(+1)$ of type $Int \rightarrow Int$ and the primitive $NOT$ of type
$Bool \rightarrow Bool$. If the requesting type is $\sigma \rightarrow
Int$, then we can only use $(+1)$, but if the requesting type is
$\sigma \rightarrow \sigma$ then we could use either $(+1)$ or
$NOT$. We define $C_\tau$ for any type $\tau$ to be the set of
primitive combinators in $C$ that are consistent with $\tau$, that is,
whose types \emph{unify} with $\tau$ (we use unification according to
the standard type theory presentation~\cite{Pierce_2002}).

Let $C_e$ be the set of primitive combinators in the leaves of
expressions $e$. We define the probability of an expression ${e \in
  \mathcal{L}}$ to be ${ p(e)~=~\prod_{c \in C_e} p(c | \tau(c)) }$
where $p(c | \tau(c))$ is the probability of using primitive $c$ when
the requesting type is $\tau(c)$. In turn, we define the conditional
probability for each combinator $c_n$ to be $p(c_n| \tau(c_n)) \propto
\frac{p}{\sum_{c_j \in C_{\tau(c_n)}} p_j}$.  That is, $p(c_n | \tau(c_n)$ is
proportional to the probability of sampling $c_n$ from the multinomial
distribution defined by the $p_n$'s, conditioned on the requesting
type, with the constant of proportionality chosen to ensure that the
sum over the probabilities of all expressions converges to a finite
value (so that the distribution over expressions is proper). For
binary trees, this is true whenever the constant of proportionality is
less than~$\frac{1}{4}$\footnote{The probability mass $M_d$ of all
  expressions of depth less than or equal to $d$ can be written with
  the recurrence relation $M_d \leq M_{d-1}^2 + M_1$. This has an
  asymptotic fixed point, as $d \rightarrow \infty$ if $x = x^2 + M_1$
  has a solution, which is true if $M_1~\leq~1/4$.}.

Now suppose that we have a set of expressions $\mathcal{E}$, and we
wish to estimate the maximum likelihood values of the $p_n$'s. If the
choice of primitive at each leaf were not conditioned on the
requesting type, we could just count all occurences of each primitive
in our expression set, and this would be proportional to the maximum
likelihood estimate. This is the ML estimator for a multinomial
distribution. However, the fact that we condition on the requesting
type makes this a considerably more difficult optimization task. One
of the simplest approximations we can make is to calculate the
frequency with which each primitive combinator appears in
$\mathcal{E}$ when the requesting type is such that it could have
been chosen. This is a straightforward calculuation, and
we use it for its convenience, finding that our empirical results
justify its use. Future inquiry will be needed to determine to what
extent a more accurate estimator can improve these results.

\subsection{Best-first enumeration of programs}
In order to explore the frontier of most promising programs, we need a
procedure that enumerates the $N$ most probable programs. There has
been recent interest in this problem, most of it focused on
enumerating the shortest program satisfying some
criterion~\cite{DBLP:conf/sfp/Katayama05}
\cite{DBLP:conf/aaip/YakushevJ09}. Our approach is to formulate the
problem as a best-first exploration of an AND/OR tree
(Figure~\ref{fig:andor})~\cite{nilsson1982principles}~\cite{DBLP:journals/cacm/Hall73}. In
words, the enumeration procedure is best described by the following
recursion: every program is either a primitive combinator or it is a
left child program applied to a right child program.

This can framed as the following AND/OR tree $\mathcal{G}$. Suppose we
want a program of type $\tau$. The root of $\mathcal{G}$ is an OR node of
type $\tau$. Its children are elements in $C_\tau$ (defined in the previous
section) and one AND node of type $\tau$ with two children. Each child of
an AND node has the same structure as the root node, with modified
types: the type of its left child is $\sigma \rightarrow \tau$ (where $\sigma$ is
a fresh type variable). Since the type of the right child is
constrained by the subtree rooted at the left child, we always expand
the left child first. Once we have a complete left child program, we
can use its target type as the type of the right child.

Recall that a valid partial solution $\mathcal{H}$ is a subtree of
$\mathcal{G}$ satisfying the following properties: the root of
$\mathcal{H}$ is the root of $\mathcal{G}$ and any OR node in
$\mathcal{H}$ has at most one child. $\mathcal{H}$ is a complete
solution (and thus a complete expression) if it is a partial solution
whose leaves are leaves of $\mathcal{G}$, if each OR node in
$\mathcal{H}$ has exactly one child, and if each AND node in
$\mathcal{H}$ is connected to all of that node's children in
$\mathcal{G}$

@@

\subsection{Re-estimating the stochastic grammar}

@@

In the formulation we consider, each task $t$ corresponds to a
polynomial $f$, and an expression $e$ solves $t$ if ${ (e\; i)=f(i) }$
for all ${ i=0,\dots,9 }$. For this experiment, the set of problems
corresponds to the set of all polynomials with degree less than or
equal to 2. In the initial library, we include the basic combinators
I, S, B, C and four arithmetic primitives, $1$, $0$, $*$, and $+$. The
choice of initial weights on these primitives can make a difference,
as it determines the relative ordering of combinators in the
enumeration step. To get a general sense of the algorithm's
performance, we set the initial weights to be slightly different on
each run, perturbing them around a uniform weighting.

\setcounter{subfigure}{0} 
\begin{figure}[ht]
\centering
\begin{subfigure}[Before]{\linewidth}
\includegraphics[width=.9\linewidth]{figures/symreg3frontiers.pdf}
\caption{Learning curves as a function of frontier size.  As frontier
  size is increased, it plateaus closer to $100\%$ performance. A
  control brute-force search over $150000$ only hits $3\%$ of the
  tasks. \label{fig:symregLearning}}
\end{subfigure}
